Antisense transcripts can quickly evolve to encode proteins

### 2023-03-10 Manuscripts/M2/antisense_genes.tex
modified:   Manuscripts/Figures/M2_main/frames.svg 	modified:   Manuscripts/Figures/M2_main/pORF_antisense_scer.svg 	modified:   Manuscripts/Figures/M2_main/pORFgain_antisense_dmelGC.svg 	modified:   Manuscripts/Figures/M2_main/pORFgain_antisense_dmelWC.svg 	modified:   Manuscripts/Figures/M2_main/pORFgain_antisense_scerGC.svg 	modified:   Manuscripts/Figures/M2_main/pORFgain_antisense_scerWC.svg 	modified:   Manuscripts/Figures/M2_main/pORFloss_antisense_dmelGC.svg 	modified:   Manuscripts/Figures/M2_main/pORFloss_antisense_dmelWC.svg 	modified:   Manuscripts/Figures/M2_main/pORFloss_antisens

We estimate the stationary and transition probabilities of antisense ORFs using the existing (sense) ORF as a reference. Specifically, we analyse the effect of the composition of the sense ORF, and of mutations in its sequence, to determine how likely it is for an antisense ORFs to exist, and to be gained or lost. Antisense ORFs can overlap with the sense ORFs in three different frames. In frame 0, the codons in the antisense ORF exactly overlap the codons in the sense ORF. In frames 1 and 2, the codons in the antisense ORF are shifted towards the 5' end of sense ORF by one and two nucleotide positions, respectively. Thus in frames 1 and 2, the sequence of an antisense codon is determined by two partially overlapping sense codons (dicodons). Due to this sequence overlap, the evolution of antisense ORFs would be constrained by the evolutionary selection pressures on the sense ORF. Furthermore, these constraints would be different for antisense ORFs located in the three different frames. We analysed the evolution of antisense ORFs when the sense ORF is under three different levels of purifying selection. The first level describes an absence of purifying selection, where any kind of mutation except a non-sense mutation (gain of stop codon) in the sense ORF is tolerated. The second level describes a weak purifying selection that allows synonymous mutations (nucleotide mutations that do not change the encoded amino acid), as well as mutations where an amino acid is substituted by a chemically similar amino acid (for example, aspartic acid to glutamic acid). Finally, in the third level that describes a strong purifying selection, only synonymous mutations are tolerated in the sense ORF.

@@

Based on these considerations, we determined the probability of finding an antisense ORF. To this end, we calculated the probability of finding a stop codon the three antisense frames, given the nucleotide composition of sense ORF including the condition that no stop codon exists in the sense ORF. Specifically, a stop codon can exist in frame 0 wherever the reverse complementary codons (\texttt{TTA},\texttt{CTA},\texttt{TCA}) exist in the sense ORF. It can exist in frames 1 and 2, overlapping with one of the 128 and 192 dicodons, respectively. We thus calculated the probability of an antisense stop codon based on the probability of the corresponding sense codons and dicodons. We calculated the probability of finding a start codon using the general GC content of the locus, ignoring any effect of the sense ORF. Using the start and stop codon probabilities, we estimated the probability of finding an antisense ORF of different lengths in each of the three frames. We did so for four different values of GC content. We found that frame 1 has the highest likelihood of harboring an antisense ORF in 99.2\% of cases. The only exceptions are ORFs shorter than 39 codons present in a DNA region with a GC content of 60\%. Even for these exceptional cases, the probability of an antisense ORF in frame 1 was no less than 91\% of the corresponding ORF probabilities in other frames.

To understand whether antisense ORFs are more likely to be found than ORFs in intergenic regions, we compared the probabilities of antisense ORFs with that of intergenic ORFs of identical GC content and lengths. We note that that the probability of intergenic ORFs is not constrained by existing ORFs and is thus purely dependent on nucleotide composition. We found that the probability of antisense ORFs in frames 0 and 2 were identical to that of the corresponding intergenic ORFs, which means that antisense ORFs in frame 1 were more likely to exist than intergenic ORFs except in the 0.8\% cases where the ORFs had less than 39 codons and a GC content of 60\%.

GC content may not accurately describe the nucleotide composition of a locus. Therefore, we also calculated the probability of antisense ORFs using actual codon and dicodon frequencies in yeast ORFs. Likewise, we calculated the probability of intergenic ORFs using the frequencies of DNA trimers in yeast intergenic genome.

>>>>

We estimate the stationary and transition probabilities of antisense ORFs using the existing (sense) ORF as a reference. Specifically, we analyse the effect of the composition of the sense ORF, and of mutations in its sequence, to determine how likely it is for an antisense ORFs to exist, and to be gained or lost. Antisense ORFs can overlap with the sense ORFs in three different frames. In frame 0, the codons in the antisense ORF exactly overlap the codons in the sense ORF. In frames 1 and 2, the codons in the antisense ORF are shifted towards the 5' end of the sense ORF by one and two nucleotide positions, respectively. Thus in frames 1 and 2, the sequence of an antisense codon is determined by two partially overlapping sense codons (dicodons). Due to this sequence overlap, the evolution of antisense ORFs would be constrained by the evolutionary selection pressures on the sense ORF. Furthermore, these constraints would be different for antisense ORFs located in the three different frames. We analysed the evolution of antisense ORFs when the sense ORF is under three different levels of purifying selection. The first level describes an absence of purifying selection, where any kind of mutation except a non-sense mutation (gain of stop codon) in the sense ORF is tolerated. The second level describes a weak purifying selection that allows synonymous mutations (nucleotide mutations that do not change the encoded amino acid), as well as mutations where an amino acid is substituted by a chemically similar amino acid (for example, aspartic acid to glutamic acid). Finally, in the third level that describes a strong purifying selection, only synonymous mutations are tolerated in the sense ORF.

@@

Based on these considerations, we determined the probability of finding an antisense ORF. To this end, we calculated the probability of finding a stop codon in the three antisense frames, given the nucleotide composition of the sense ORF, including the condition that no stop codon exists in the sense ORF. Specifically, a stop codon can exist in frame 0 wherever the reverse complementary codons (\texttt{TTA},\texttt{CTA},\texttt{TCA}) exist in the sense ORF. It can exist in frames 1 and 2, overlapping with one of the 128 and 192 dicodons in the sense ORF, respectively. We next calculated the probability of an antisense stop codon based on the probability of the corresponding sense codons and dicodons. We calculated the probability of finding a start codon using the general GC content of the locus, ignoring any effect of the sense ORF. Using the start and stop codon probabilities, we estimated the probability of finding an antisense ORF of different lengths in each of the three frames. We did so for four different values of GC content. We found that frame 1 has the highest likelihood of harboring an antisense ORF in 99.2\% of cases. The only exceptions are ORFs shorter than 39 codons present in a DNA region with a GC content of 60\%. Even for these exceptional cases, the probability of an antisense ORF in frame 1 is no less than 91\% of the corresponding ORF probabilities in the other frames.

To understand whether antisense ORFs are more likely to be found than ORFs in intergenic regions, we compared the probabilities of antisense ORFs with that of intergenic ORFs of identical GC content and lengths. We note that that the probability of intergenic ORFs is not constrained by existing ORFs and is thus purely dependent on nucleotide composition. We found that the probability of antisense ORFs in frames 0 and 2 were identical to that of the corresponding intergenic ORFs, which means that antisense ORFs in frame 1 are more likely to exist than intergenic ORFs except in the 0.8\% cases where the ORFs have a GC content of 60\% and less than 39 codons.

GC content may not accurately describe the nucleotide composition of a locus. Therefore, we also calculated the probability of antisense ORFs using actual codon and dicodon frequencies in yeast ORFs. Likewise, we calculated the probability of intergenic ORFs using the frequencies of DNA trimers in yeast intergenic genome. With this analysis, we found that antisense ORFs longer than 54 and 75 codons in frames 1 and 2, respectively, are more likely to occur than intergenic ORF of the same lengths. For all ORF lengths, antisense ORFs in frame 3 are less likely to occur than intergenic ORFs.
